Fail Faster

### 2025-03-21 paper/paper.tex

We lay a path forward.
First, we show how to use \textit{multi-stage programming}---a classic technique for 
building fast DSLs “without regret”~\cn{}---to eliminate the overhead of 
many common generator abstractions.
Second, we demostrate that entropy source is an undervalued consideration in generator library design
by isolating its performance contribution, and showing that significant
speedups are possible.
These insights culminate in \name, a recipe for building generator libraries that marry high-level design
with low-level performance.

@@

To evaluate our technique, we implement staged PBT libraries in two strict functional languages: OCaml and Scala 3.
In OCaml, we build \camlname{}, based on \bq{}---a
high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street \cn{}.
In Scala 3, we build \scalaname{}, based on ScalaCheck---a very
mature PBT library and the {\em de facto} choice\bcp{the wording here
  implies that there are other choices in Scala, some of which might
  be reasonable, and tht base\_quickcheck is not the de facto choice
  in OCaml...} in Scala \cn{}.
In both languages, our libraries are direct drop-in
replacements\bcp{that's not quite accurate, is it? I mean, they have
  the same functionality, but can you take a generator that works with
base\_quickcheck and just link it against our version without changing
the generator itself? That's what ``drop-in replacement'' means to me.} that are
\textit{100\% semantically equivalent}: that is, given identical random seeds, a generator written with our technique
produces exactly the same sequence of values as the same generator written with the original generator library.
The semantic equivalence between the original generators and \name{} generators allows us to
perform clean, fine-grained comparisons, as differences in generation speed are
solely attributable to the staging and \rand optimizations.
\jwc{BCP thinks this should be a little stronger: ``and not obscured by the probablistic nature of generation'', or something like that. ``Comparing distributions of speed is hard, comparing pointwise is easier.''. This needs to have its own sentence here.}

>>>>

We lay a path forward.
First, we show how to use \textit{multi-stage programming}---a classic technique for 
building fast DSLs “without regret”~\cn{}---to eliminate the overhead of 
many common generator abstractions.
Second, we demostrate that entropy source is an undervalued consideration in generator library design
by isolating its performance contribution, and showing that significant
speedups are possible---even while maintaining output equivalence.
These insights culminate in \name, a recipe for building generator libraries that marry high-level design
with low-level performance.

@@

To evaluate our technique, we implement staged PBT libraries in two strict functional languages: OCaml and Scala 3.
In OCaml, we build \camlname{}, based on \bq{}---a
high-quality PBT library in OCaml, authored by software engineers at the trading firm Jane Street \cn{}.
In Scala 3, we build \scalaname{}, based on ScalaCheck---the language's standard PBT library, and by far the most widely-used\cn{}.
In both languages, our libraries provide identical functionality and behavior: given the same random seed, a generator written with our technique produces exactly the same sequence of values as an equivalent generator written with the original library.
This semantic equivalence enables pointwise, rather than distributional, comparisons, ensuring that performance differences are attributable \textit{solely} to staging and \rand optimizations.
% \jwc{BCP thinks this should be a little stronger: ``and not obscured by the probablistic nature of generation'', or something like that. ``Comparing distributions of speed is hard, comparing pointwise is easier.''. This needs to have its own sentence here.}
